Re-run, Repeat, Reproduce, Reuse, Replicate:\\Transforming Code into Scientific Contributions

### 2017-08-20 draft.tex
Improving the replicable section. Still needs some love.

% 1. parameters and 2. code version with results
This is why results files should come alongside their context, i.e. an exhaustive list of the parameters used as well as a precise description of the execution environment, as the R$^3$ code does. The code itself is part of that context: the version of the code must be recorded as part of the provenance data. It is common for different results or different figures to have been generated by different versions of the code. Ideally, all results should originate from the same (and last) version of the code. But for long or expensive computations, this may not be feasible. In that case, the result files should contain the version of the code that was used to produce it. This information can be obtained from the version control software. This also allows, if some errors are found and corrected after some results have been obtained, to identify which ones should be recomputed. In R$^3$, the code records the current git hash, and whether the repository holds uncommited changes when the computation starts.\\

% scripts to generates figures & data behind graphs
The code should also contain all the scripts that were used to produce the figures. Plotting code may seem mundane, but it is as vulnerable as any other piece of the code to bugs and errors. With a code that is re-runnable, repeatable and contains such scripts, a user can quickly check that published figures can be reproduced. However, figures by themselves are imprecise and cumbersome, and sometimes plain unusable, when it comes to checking that the reproduced data match the one published in the article. A published figure should always be accompanied by its underlying data to allow straightforward numeric comparisons.\\

@@

% reproducibility implies availability
It should be obvious by now that \emph{reproducibility implies availability}. As shown in \citep{Collberg:2016}, code is often unavailable, or only available upon request. While the latter may seem sufficient, changes in email address, changes in career, retirement, a busy inbox or poor archiving practices can make a code just as unreachable. Code \emph{and} input \emph{and} result files \emph{and} figure data should be available with the published article, as supplementary data, or through a DOI link to a scientific repository such as Figshare or Zenodo\footnote{Online code repositories such as GitHub \emph{are not} scientific repositories, and may disappear, change name or change their access policy at any moment. Direct links to them are not perpetual, and, when used, they should always be supplemented by a DOI link to a scientific archive.}.\\

@@

% Summary
To recap, reproducibility implies re-runnability and repeatability and availability, yet imposes additional conditions. Dependencies and platforms must be described as precisely and a specifically as possible. Parameters values and inputs should accompany the result files. The data and scripts behind the graphs must be published. Unit tests are a good way to embed self-diagnostics of reproducibility in the code. Reproducibility is hard, yet tremendously necessary.

@@

For you, first. Because the you now and the you in two years are two different persons. Details on how to use the code, its limitations, its quirks, may be present to your mind now, but will probably escape you in six months \parencite{Donoho:2009}. Here, comments and documentation can make a significant difference. In fact, a source code reflects the results of all the decisions that were made to create it, but not the reasons behind those decisions. In science, where the method and its justification matter as much as the results, those reasons are precious knowledge. In that context, a comment on how a given parameter was chosen (optimization, experimental data, educated guess), why a library was chosen over another (conceptual or technical reasons?) is valuable information.\\

@@

Despite all this, reusability is often overlooked, and it is not hard to see why. Scientists are rarely trained in software engineering, and reusability can represent an expensive endeavour if undertaken as an afterthought, for little tangible short-term benefits, for a codebase that might, after all, see only a single use. And, in fact, reusability is not as indispensable a requirement as the other ones presented here. Yet, some simple measures can tremendously increase reusability, and at the same time strengthen reproducibility and re-runnability.\\

@@

Documentation, in fact, is one of the most potent tools for reusability. A proper documentation on how to install and run the software often makes the difference for other researchers whether they manage to use the code or not.\\

Furthermore, a comment describing what each function does, however evident, can avoid hours of head-scratching. Great code may need few comments. Scientists, however, are not always brilliant developers. Code quality, in science, averages on mediocre. Of course, bad, complicated code should be rewritten until is simple enough to explain itself. But realistically, this is hardly ever going to be done: there is simply not enough incentive for it. There, a comment that explains the intention and scientific reasons behind a block of code can tremendously useful.\\

@@

Re-usability is not an absolute requirement for scientific code. But it has many benefits, and a few simple measures can foster it considerably.

@@

% reusable is not foolproof
Having made a software reusable offers an additional way to find bugs, especially if your scientific contribution is popular. Unfortunately, this is not always effective, and some recent cases have dramatically illustrated the tremendous impact a bug can have in science
\citep{Eklund:2016} or in our every-day life \citep{Durumeric:2014}. This is
why, as explained by Peng et al. \cite{Peng:2006}, {\em the replication of
  important findings by multiple independent investigators is fundamental to
  the accumulation of scientific evidence}.\\

@@

% conceptual description != replicable description
This is because most papers strive to communicate the main ideas behind their contribution is terms as simple and as clear as possible, so that the reader may be able to easily understand them and the results that are presented to him. Trying to ensure replicability in the main text adds a myriad of esoteric details that are not conceptually significant, and clutter the explanations. Therefore, unless the writer dedicates an addendum or a section of the supplementary information for technical details specifically aimed at replicability, the information will not be there because there are incentives not to do so.\\

% errors may always remain: reproducibility does not make reproducibility useless.
But even when those details are present, the best efforts may fall short because
an oversight, a typo or a difference between what is evident for the writer
and for the reader \citep{Mesnard:2016}. In that case, a reproducible code is a crucial backup. Hence,
replicability does not negate the necessity of reproducibility. To illustrate
this, let us consider what could be the textual description of the random
walker (as it would be written in an article describing it):
%
\begin{quotation}
{\em The model uses the Mersene Twister generator initialized with the seed 0. At each iteration, a uniform number between -1 (included) and +1 (excluded) is drawn and the sign of the result is used for generating a positive or negative step.}
\end{quotation}
%
This description, while somewhat precise, forgoes---as it is common---the initialization of the variables (here the starting value of the walk: {\tt 0}), and the technical details about which implementation of the RNG is used.\\

% NumPy and Python: differences in the Mersene Twister
It may look innocuous. After all, the
\href{https://docs.python.org/3.6/library/random.html}{Python documentation},
states that {\em Python uses the Mersenne Twister as the core generator. It produces
53-bit precision floats and has a period of 2**19937-1.} Someone trying to replicate the work however might choose to use the RNG from the \href{http://www.numpy.org/}{NumPy library}.  The NumPy library is extensively used in the science community, and it provides an implementation of the Mersene Twister generator too. Unfortunately, the way the seed is interpreted by the two implementations is different, yielding different random sequences.\\

>>>>

% 1. parameters and 2. code version with results
This is why results files should come alongside their context, i.e. an exhaustive list of the parameters used as well as a precise description of the execution environment, as the R$^3$ code does. The code itself is part of that context: the version of the code must be recorded as part of the provenance data. It is common for different results or different figures to have been generated by different versions of the code. Ideally, all results should originate from the same (and last) version of the code. But for long or expensive computations, this may not be feasible. In that case, the result files should contain the version of the code that was used to produce it. This information can be obtained from the version control software. This also allows, if some errors are found and corrected after some results have been obtained, to identify which ones should be recomputed. In R$^3$, the code records the last git commit hash, and whether the repository holds uncommited changes when the computation starts.\\

% scripts to generates figures & data behind graphs
Published results should come from version of the code where every file has been commited. This includes pre-processing, post-processing and plotting code. Plotting code may seem mundane, but it is as vulnerable as any other piece of the code to bugs and errors. With a code that is re-runnable, repeatable and contains plotting scripts, a user can quickly reproduce published figures. When it comes to checking that the reproduced data match the one published in the article, however, figures can reveal themselves to be imprecise and cumbersome, and sometimes plain unusable. Ideally, published figures should be accompanied by their underlying data (coordinates of the plotted points) in the supplementary data to allow straightforward numeric comparisons.\\

@@

% reproducibility implies availability
It should be obvious by now that \emph{reproducibility implies availability}. As shown in \citep{Collberg:2016}, code is often unavailable, or only available upon request. While the latter may seem sufficient, changes in email address, changes in career, retirement, a busy inbox or poor archiving practices can make a code just as unreachable. Code \emph{and} input data \emph{and} result data should be available with the published article, as supplementary data, or through a DOI link to a scientific repository such as Figshare or Zenodo\footnote{Online code repositories such as GitHub \emph{are not} scientific repositories, and may disappear, change name or change their access policy at any moment. Direct links to them are not perpetual, and, when used, they should always be supplemented by a DOI link to a scientific archive.}.\\

@@

% Summary
To recap, reproducibility implies re-runnability and repeatability and availability, yet imposes additional conditions. Dependencies and platforms must be described as precisely and as specifically as possible. Parameters values and inputs should accompany the result files. The data and scripts behind the graphs must be published. Unit tests are a good way to embed self-diagnostics of reproducibility in the code. Reproducibility is hard, yet tremendously necessary.

@@

For you, first. Because the you now and the you in two years are two different persons. Details on how to use the code, its limitations, its quirks, may be present to your mind now, but will probably escape you in six months \parencite{Donoho:2009}. Here, comments and documentation can make a significant difference. Source code reflect the results of the decisions that were made during its creation, but not the reasons behind those decisions. In science, where the method and its justification matter as much as the results, those reasons are precious knowledge. In that context, a comment on how a given parameter was chosen (optimization, experimental data, educated guess), why a library was chosen over another (conceptual or technical reasons?) is valuable information.\\

% Research team typically experience a high turnover of interns, doctoral and post-doctoral students. Can you reproduce the results of member after leave your team?

@@

Despite all this, reusability is often overlooked, and it is not hard to see why. Scientists are rarely trained in software engineering, and reusability can represent an expensive endeavour if undertaken as an afterthought, for little tangible short-term benefits, for a codebase that might, after all, see only a single use. And, in fact, reusability is not as indispensable a requirement as the other ones presented here. Yet, some simple measures can tremendously increase reusability, and at the same time strengthen reproducibility and re-runnability over the long-term.\\

@@

@@

Documentation is one of the most potent tools for reusability. A proper documentation on how to install and run the software often makes the difference whether other researchers manage to use the code or not. A comment describing what each function does, however evident, can avoid hours of head-scratching. Great code may need few comments. Scientists, however, are not always brilliant developers. Code quality, in science, averages on mediocre. Of course, bad, complicated code should be rewritten until is simple enough to explain itself. But realistically, this is not always going to be done: there is simply not enough incentive for it. There, a comment that explains the intention and scientific reasons behind a block of code can tremendously useful.\\

Reusability is not a strict requirement for scientific code. But it has many benefits, and a few simple measures can foster it considerably.

% But a few good practices, observed during the development of the code can have significant benefits for you, your team and any other reasearche desirous to use your code.

@@

% reusable is not foolproof
Having made a software reusable offers an additional way to find errors, especially if your scientific contribution is popular. Unfortunately, this is not always effective, and some recent cases have dramatically illustrated the tremendous impact a bug can have in science
\citep{Eklund:2016} or in our every-day life \citep{Durumeric:2014}. This is
why, as explained by Peng et al. \cite{Peng:2006}, {\em the replication of
  important findings by multiple independent investigators is fundamental to
  the accumulation of scientific evidence}.\\

@@

% conceptual description != replicable description
This is because most papers strive to communicate the main ideas behind their contribution is terms as simple and as clear as possible, so that the reader may be able to easily understand them and the results that are presented to him. Trying to ensure replicability in the main text adds a myriad of esoteric details that are not conceptually significant and clutter the explanations. Therefore, unless the writer dedicates an addendum or a section of the supplementary information for technical details specifically aimed at replicability, the information will not be there because there are incentives not to do so.\\

% errors may always remain: reproducibility does not make reproducibility useless.
But even when those details are present, the best efforts may fall short because an oversight, a typo or a difference between what is evident for the writer and for the reader \citep{Mesnard:2016}. Minute changes in the numerical estimation of a common first-order differential equation can have significant impact \citep{Crook:2013}. A reproducible code plays a role that cannot be adequately fulfilled by the conceptual description of it found in the article. Hence, replicability does not negate the necessity of reproducibility. To illustrate this, let us consider what could be the textual description of the random walker (as it would be written in an article describing it):
%
\begin{quotation}
{\em The model uses the Mersene Twister generator initialized with the seed 0. At each iteration, a uniform number between -1 (included) and +1 (excluded) is drawn and the sign of the result is used for generating a positive or negative step.}
\end{quotation}
%
This description, while somewhat precise, forgoes---as it is common---the initialization of the variables (here the starting value of the walk: {\tt 0}), and the technical details about which implementation of the RNG is used.\\

% NumPy and Python: differences in the Mersene Twister
It may look innocuous. After all, the
\href{https://docs.python.org/3.6/library/random.html}{Python documentation},
states that "Python uses the Mersenne Twister as the core generator. It produces
53-bit precision floats and has a period of 2**19937-1". Someone trying to replicate the work however might choose to use the RNG from the \href{http://www.numpy.org/}{NumPy library}.  The NumPy library is extensively used in the science community, and it provides an implementation of the Mersene Twister generator too. Unfortunately, the way the seed is interpreted by the two implementations is different, yielding different random sequences.\\
